Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-16 paper/main.tex
working through section 3

For both $\Sigma_S$ and $\Sigma_W$, we use Mat\`ern kernels with unit scale and
unique lengthscales,
\[
\begin{array}{l}
\Sigma_S(W)^{i,j} = k\left(||w_i - w_j||^2, \theta_S\right) \\[7pt]
\Sigma_W(X)^{i,j} = k\left(||x_i - x_j||^2, \theta_W\right)
\end{array}
\quad\textrm{where}\quad
k(d, \theta) = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta}} + 
  \frac{5d^2}{3\theta}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta}}\right).
\]
In our one-dimensional setting, we embrace unit scale on $S$ to preserve identifiability
with $\theta_S$ \textcolor{red}{(@Steve - can you cite that paper here?)}.
Unit scale on latent $W$ was recommended by \citet{sauer2023active} for similar reasons.

@@

Our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, but this requires inferring latent quantity $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~(\ref{eq:dgphm}) into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\mathbf{0}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
Details are provided in \ref{sec:apdx_int_lik}. The incorporation of $\Sigma_\epsilon$ in the
Gaussian likelihood of this outer layer is an essential upgrade to previous DGP inferential schemes.
This covariance incorporates knowledge of the smoothly varying dynamics from which $\bar{Y}$ was
generated.

We embrace fully-Bayesian inference of latent $W$ and kernel hyperparameters $\{\theta_S, \theta_W\}$.
Specifically, we use elliptical slice sampling \citep[ESS;][]{murray2010elliptical} to 
generate posterior samples of $W$, using Eq.~(\ref{eq:prior}) for proposal generation and 
Eq.~(\ref{eq:likelihood}) for likelihood-based acceptance.  
Under a reference prior $\pi(\tau^2)\propto\frac{1}{\tau^2}$,
we integrate $\tau^2$ from the likelihood of Eq.~(\ref{eq:likelihood}).  The resulting marginal 
posterior is equivalent to that with the plug-in estimator 
$\hat{\tau}^2 = \dots$ \citep[][Chapter 5]{gramacy2020surrogates}.
\textcolor{red}{(With $\Sigma_\epsilon$ on there, we can't actually do this.. unless we divide
$\Sigma_\epsilon$ by $\tau^2$...)}.
For $\theta_S$ and $\theta_W$, after pre-scaling, we adopt the default prior specifications 
provided in the {\tt deepgp} {\sf R}-package \citep{deepgp}.
We then integrate Metropolis-Hastings sampling of these parameters in a Gibbs framework 
with the ESS sampling of $W$ \citep{sauer2023active}.

>>>>

For both $\Sigma_S$ and $\Sigma_W$, we use Mat\`ern kernels with unit scale and
unique lengthscales,
\[
\begin{array}{l}
\Sigma_S(W)^{i,j} = k\left(||w_i - w_j||^2, \theta_S\right) \\[7pt]
\Sigma_W(X)^{i,j} = k\left(||x_i - x_j||^2, \theta_W\right)
\end{array}
\quad\textrm{where}\quad
k(d, \theta) = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta}} + 
  \frac{5d^2}{3\theta}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta}}\right).
\]
In our one-dimensional setting, we embrace unit scale on $S$ to preserve parsimony 
and identifiability with $\theta_S$ \textcolor{red}{(@Steve - can you cite that paper here?)}.
Unit scale on latent $W$ was recommended by \citet{sauer2023active} for similar reasons.

@@

Our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, but this requires inferring latent quantity $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~(\ref{eq:dgphm}) into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\mathbf{0}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{monoGP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
Details, including generalizations for $\boldsymbol{\mu}_S\neq\mathbf{0}$, are provided in \ref{sec:apdx_int_lik}. 
The incorporation of $\Sigma_\epsilon$ in the
Gaussian likelihood of this outer layer is an essential upgrade to previous DGP inferential schemes.
This covariance incorporates knowledge of the smoothly varying dynamics from which $\bar{Y}$ was
generated.

We embrace fully-Bayesian inference of latent $W$ and kernel hyperparameters $\{\theta_S, \theta_W\}$.
Specifically, we use elliptical slice sampling \citep[ESS;][]{murray2010elliptical} to 
generate posterior samples of $W$, using Eq.~(\ref{eq:prior}) for proposal generation and 
Eq.~(\ref{eq:likelihood}) for likelihood-based acceptance.  
For $\theta_S$ and $\theta_W$, after pre-scaling, we adopt the default prior specifications 
provided in the {\tt deepgp} {\sf R}-package \citep{deepgp}.
We then integrate Metropolis-Hastings sampling of these parameters in a Gibbs framework 
with the ESS sampling of $W$ \citep{sauer2023active}.
